---
title: "03 Exploratory-plots"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
    smooth_scroll: yes
  pdf_document:
    toc: yes
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
suppressPackageStartupMessages({
library(data.table)
library(ggplot2)
library(RColorBrewer)
library(ggrepel)
library(GenomicRanges)
library(stringr)
library(RColorBrewer)
library(rtracklayer)
library(dplyr)
library(ComplexHeatmap)
library(viridis)
library(bsseq)
library(hrbrthemes)
library(limma)
source("utils.R")})

knitr::opts_chunk$set(warning=FALSE, 
                      message=FALSE,
                      fig.width=12, fig.height=7)
```

# Exploratory Patient Plots

Load the previously processed files (01_data_processing_split.R)
```{r}
bismarkBSseqTrain <- readRDS("./data/bismarkBSseq_train_ids.rds")
meta <- pData(bismarkBSseqTrain)

# Probe meta data
probAnnot <- readRDS("./meta_data/probeAnnot.rds")
probeRegions <- makeGRangesFromDataFrame(as.data.frame(probAnnot), 
                                         keep.extra.columns=TRUE)

# Get methylation values
# per Probe
methProbes <- as.data.table(bsseq::getMeth(bismarkBSseqTrain, 
                                           regions=probeRegions, 
                                           type="raw", what="perRegion"))
colnames(methProbes) <- as.character(colData(bismarkBSseqTrain)$sample)

probAnnot[grepl("cADN", methylated_in) & grepl("SSL", methylated_in),]$methylated_in="cADN and SSL specific markers"
probAnnot$methylated_in <- factor(probAnnot$methylated_in, 
                                  levels=c("Control regions",
                                           "Age dependent markers",
                                           "cADN markers",
                                           "SSL markers",
                                           "cADN and SSL specific markers"), 
                                  ordered=TRUE)

methProbes$category <- probAnnot$methylated_in
methProbes$probe <- probAnnot$probe

# per CpG
methFracCpG <- bsseq::getMeth(bismarkBSseqTrain, type="raw", withDimnames=TRUE)
colnames(methFracCpG) <- colData(bismarkBSseqTrain)$sample
```

## scater plots for all patients
```{r}
trainMeta<- readRDS("./data/train_samples_ids.rds")
trainSamples <- trainMeta$sample
trainPatients <- unique(trainMeta$patient)
```

```{r, eval=FALSE, echo=FALSE}
for(i in trainPatients){
  samps <- as.character(meta$sample[meta$patient_id == i])
  p <- ggplot(methProbes,aes_string(x=samps[2], y=samps[1], color = "category")) + 
    geom_point() +
    theme_bw() +
    xlab(sprintf("Methylation level %s",samps[2])) +
    ylab(sprintf("Methylation level %s",samps[1]))
    #ggtitle("Normal Vs tumor methylation per probe")+
    theme_ipsum()
  ggsave(file.path("./plots", 
                   sprintf("normalVstumor_%s.png",i)), p, width = 10, height = 5)
}
```

# MDS Plot

MDS plot of all train samples
```{r}
# Remove probes that are "Age dependent DMRs" and "tumors and normal mucosa"
methProbesSub <- methProbes[!methProbes$category %in% c("Age dependent markers", 
                                                        "control regions"),]

# Choose only methylation values from this table
methProbesSub <- methProbesSub[,c(as.character(trainSamples)), with=FALSE]
dim(methProbesSub)

# Transform methylation values for the MDS plot
methsTR <- asin(2*as.matrix(methProbesSub)-1)

# Calculate components for 2-dimension MDS plot
#as a rule of thumb, I choose the top 10% of the total
m2 <- limma::plotMDS(methsTR, dim.plot=c(1,2), top = 100, labels = colnames(methsTR))[c("x", "y")]
m2 <- as.data.frame(m2)
colnames(m2) <- c("MDS1","MDS2")
m2 <- as.data.frame(m2)
head(m2)

# Add tissue and sample name to information to m2
m2$tissue <- trainMeta$condition
m2$sample <- trainMeta$sample
head(m2)

# Plot MDS
ggplot(m2) + 
  geom_point(aes(x = MDS1, y=MDS2, color = tissue)) +
  geom_label_repel(aes(x = MDS1, y=MDS2, label = sample)) +
  theme_bw() +
  xlab("MDS 1") +
  ylab("MDS 2") +
  ggtitle("MDS plot with top 100 probes with largest standard deviations")
ggsave("./plots/mds_labelled_points_ids_train.svg")
ggsave("./plots/mds_labelled_points_ids_train.png")

ggplot(m2) + 
  geom_point(aes(x = MDS1, y=MDS2, color = tissue), size=2, alpha=0.8) +
  scale_color_brewer(palette="Set2")+
  theme_minimal() +
  labs(title="MDS",
       subtitle="top 100 probes with largest standard deviations",
       x="MDS 1",
       y="MDS 2")
ggsave("./plots/mds_points_ids_train.svg")
ggsave("./plots/mds_points_ids_train.png")
```

# Exploratory Probe Plots

Number of CpGs per probe
```{r}
cov2 <- getCoverage(bismarkBSseqTrain, type = "Cov")
cpgAnnot <- rowData(bismarkBSseqTrain)

cov2 <- data.frame(cov2)
probeCov <- cbind(cpgAnnot, "mean_cov"=rowMeans(cov2))
probeCov <- as.data.table(probeCov)

# Get number of CpGs per probe
probeNCpGs <- probeCov[,.(n_cpgs=.N,
                          mean_cov=mean(mean_cov)), 
                          by=c("probe")]

ggplot(probeNCpGs, aes(x=n_cpgs))+
geom_histogram(color="black", fill="white")+
xlab("N CpGs")+
ylab("Probe count")+
theme_bw()

probeNCpGs$set <- "all Probes"
ggplot(probeNCpGs, aes(x=set, y=n_cpgs))+
geom_violin()+
geom_boxplot(color="black", fill="white", outlier.size=0.7)+
xlab("")+
ylab("N CpGs per Probe")+
theme_bw()

# n CpGs vs Coverage
ggplot(probeNCpGs, aes(x=n_cpgs, y=mean_cov))+
geom_point(size=0.7)+
geom_density2d()+
xlab("N CpGs")+
ylab("Mean probe coverage")+
theme_bw()
```

Calculate standard deviation and methylation differences per probe
```{r, warning=FALSE}
# bring table in long format
methProbeTrainTable <- as.data.table(methProbes)
methProbeTrainTable <- melt(methProbeTrainTable,
                            id.vars=c("probe", "category"))

names(methProbeTrainTable) <- c("probe", "category", 
                                "sample", "mean_methylation")
methProbeTrainTable$sample <- as.character(methProbeTrainTable$sample)

# get condition/tissue (cancer vs normal mucosa)
methProbeTrainTable[,c("patient", "condition"):=tstrsplit(sample, "_")]
methProbeTrainTable[,tissue:=fifelse(condition=="NM", "NM", "CC")]

# calculate standard deviation per condition/tissue and probe
methProbeTrainTable[,sd_methylation:=sd(mean_methylation, na.rm=TRUE),
                    by=c("tissue", "probe")]

# calculate mean sd across condition/tissue
methProbeTrainTable[,mean_sd_methylation:=mean(sd_methylation), by=c("probe")]

# get cancer probes
methProbeCCTrainTable <- subset(methProbeTrainTable, tissue=="CC")
setorder(methProbeCCTrainTable, patient, probe)

# get normal probes
methProbeNMTrainTable <- subset(methProbeTrainTable, tissue=="NM")
setorder(methProbeNMTrainTable, patient, probe)

# calculate the methylation difference
methDiff <- methProbeCCTrainTable$mean_methylation-methProbeNMTrainTable$mean_methylation
methDiffTable <- data.table(meth_diff=methDiff,
                            probe=methProbeNMTrainTable$probe,
                            patient=methProbeCCTrainTable$patient,
                            condition=methProbeCCTrainTable$condition,
                            category=methProbeCCTrainTable$category,
                            mean_sd_methylation_cancer=methProbeCCTrainTable$sd_methylation)
                            #mean_sd_methylation_normal=methProbeNMTrainTable$sd_methylation,
                            #methylation_cancer=methProbeCCTrainTable$mean_methylation,
                            #methylation_normal=methProbeNMTrainTable$mean_methylation)

methProbeTrainTable <- merge(methProbeTrainTable, 
                             methDiffTable[,c("meth_diff", "patient", "probe"),
                                              #"mean_sd_methylation_cancer", "mean_sd_methylation_normal"), 
                                             with=FALSE], 
                             by=c("patient", "probe"))
methProbeTrainTable[,median_methylation_diff:=median(meth_diff, na.rm=TRUE),
                    by=c("probe")]
```

Plot SD of probes across patients. SD calculated per condition (cancer/NM) and both SDs are averaged per probe.
```{r}
ggplot(methProbeTrainTable, aes(x=median_methylation_diff, 
                                y=mean_sd_methylation,
                                color=category))+
geom_point(size=0.01)+
theme_bw()+
xlab("Methylation Difference")+
ylab("Methylation SD (per condition)")+
guides(colour = guide_legend(override.aes = list(size=2)))+
theme_ipsum()
ggsave("./plots/probe_differences_sd.png", width=8, height=5)

#methProbeTrainTable[,tissue:=fifelse(tissue=="CC", "adenoma", "nm")]
ggplot(methProbeTrainTable, aes(x=median_methylation_diff, 
                                y=sd_methylation,
                                color=category))+
geom_point(size=0.01)+
theme_bw()+
facet_grid(cols=vars(tissue))+
xlab("Methylation Difference")+
ylab("Methylation SD (per condition)")+
guides(colour = guide_legend(override.aes = list(size=2)))+
theme_ipsum()
ggsave("./plots/probe_differences_sd_facetted.png", width=12, height=5)
```

## Heatmap 

Patient x probes heatmap
```{r}
# Read metadata
meta <- subset(meta, sample %in% trainSamples)
meta$condition <- factor(meta$condition, c("NM","cADN","SSL"))

allCategories <- unique(methProbes$category)
drawHm(meta, methProbes, allCategories)
svg(filename="./plots/methylation_level_hm_train.svg", width = 10, height = 15)
dev.off()
```

Heatmap of methylation differences per probe x patient. Patients ordered by mean methylation difference.
Probes ordered by sd within the cancer sample.
```{r}
setorder(methDiffTable, condition, mean_sd_methylation_cancer)
patients <- unique(methDiffTable$patient)
meanDiffPatient <- methDiffTable[,.(mean_methylation_patient=mean(meth_diff, na.rm=TRUE)), 
                                 by=c("patient")]$mean_methylation_patient
ord <- order(meanDiffPatient)

methDiffTable$patient <- factor(methDiffTable$patient, 
                                levels=patients[ord])

ggplot(methDiffTable, aes(patient, as.factor(probe), fill=meth_diff)) + 
                      geom_tile(colour = "black", size=0.01)+
    scale_fill_viridis(option="inferno",
                       alpha=0.9,
                       guide=guide_colourbar(title="Difference in mean methylation"))+
    facet_grid(rows=vars(category), cols=vars(condition), 
               scales="free", space="free_y")+
    scale_x_discrete(position = "bottom")+
    ylab("Probes")+
    theme_minimal()+
    theme(axis.text.y = element_blank(),
          axis.text.x = element_text(angle=60, hjust=1),
          axis.title.x = element_blank(),
          panel.spacing = unit(0.5, "lines"),
          strip.background.x=element_rect(color="black", fill=NA, linetype="solid"),
          #strip.background.y=element_rect(color="black", fill=NA, linetype="solid"),
          #axis.title.y =element_blank(),
          strip.text.y = element_text(angle = 360, face="bold"),
          strip.text.x=element_text(face="bold"))

ggsave("./plots/heatmap_meth_diff_train.svg", units=c("cm"),
       width=30, height=20)
```

The samples S3, S5, A3 show very little difference on the probe level, hence they are here compared on the per CpG-level
```{r, outliers}
methOutliers <- methFracCpG[, c(grep("S5", colnames(methFracCpG), fixed=TRUE), 
                                grep("S3", colnames(methFracCpG), fixed=TRUE),
                                grep("A3_NM", colnames(methFracCpG), fixed=TRUE),
                                grep("A3_cADN", colnames(methFracCpG), fixed=TRUE))]
methOutliers <- as.data.table(methOutliers)
methOutliers$CpG <- 1:nrow(methOutliers)
methOutliers <- melt(methOutliers, id.vars=c("CpG"))
setnames(methOutliers, c("variable", "value"), c("sample", "meth_fraction"))
methOutliers[,patient:=tstrsplit(sample, split="_", keep=1)]

outlierPlot <- ggplot(methOutliers, aes(as.factor(CpG), sample, fill=meth_fraction)) + 
                      geom_tile(size=0)+
    scale_fill_viridis(option="inferno",
                       alpha=0.9,
                       guide=guide_colourbar(title="Methylated Fraction"))+
    facet_grid(rows=vars(patient), scales="free", space="free_y")+
    scale_x_discrete(position = "bottom")+
    ylab("Samples")+
    theme_minimal()+
    theme(#axis.text.y = element_blank(),
          axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          panel.spacing = unit(0.5, "lines"),
          strip.background.x=element_rect(color="black", fill=NA, linetype="solid"),
          #strip.background.y=element_rect(color="black", fill=NA, linetype="solid"),
          #axis.title.y =element_blank(),
          strip.text.y = element_text(angle = 360, face="bold"),
          strip.text.x=element_text(face="bold"))

ggsave("./plots/outliersPerCpG.svg", units=c("cm"),
       width=20, height=15)
```

# Session Information

```{r, sessioninfo}
sessionInfo()
```

