---
title: "07 Validation-plots"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
    smooth_scroll: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
library(bsseq)
library(data.table)
library(FDb.InfiniumMethylation.hg19)
library(rtracklayer)
library(GEOquery)
library(ggplot2)
library(ggrepel)
library(hrbrthemes)
library(viridis)
library(RColorBrewer)
source("utils.R")})

set.seed(1896)
knitr::opts_chunk$set(warning=FALSE, 
                      message=FALSE,
                      fig.width=12, fig.height=7)
```

load probe-level statistics
```{r, select top probes}
resultsCpG <- readRDS("./data/dmTestCpGs.rds")
resultsProbes <- readRDS("./data/dmTestProbes.rds")
topProbes <- readRDS("./data/topProbes.rds")
probeAnnot <- readRDS("./meta_data/probeAnnot.rds")
```

Volcano Plot:
```{r, volcano plot}
#resultsProbes <- merge(resultsProbes, 
#                       probeAnnot, 
#                       by.x=c("master_id"), by.y=c("master_id"),
#                       all.x=TRUE, all.y=FALSE)

volcanoPlot <- ggplot(resultsProbes, aes(x=log_fc, y=-log10(fdr) , color=methylated_in))+
               geom_point(size=2)+
               geom_label_repel(data=subset(resultsProbes, probe_id %in% topProbes$probe_id),
                                aes(x=log_fc, y=-log10(fdr), label=probe_id), max.overlaps = Inf,
                                key_glyph = "point")+
               geom_vline(xintercept=1, linetype="dashed", lwd=0.5, color="red")+
               geom_vline(xintercept=-1, linetype="dashed", lwd=0.5, color="red")+
               geom_hline(yintercept=-log10(0.05), linetype="dashed", lwd=0.5, color="black")+
               scale_color_brewer(palette="Pastel1", name="methylated in")+
               labs(x="Log FC",
                    y="-log10 FDR")+
               theme_ipsum()
volcanoPlot

ggsave(file="./plots/volcanoPlotTrain.png", volcanoPlot, width=25, height=15, units=c("cm"))
```

# Validation data

## GSE131013

See script `gse131013_process.Rmd` for processing.
```{r, load GSE131013 data}
gse131013 <- readRDS("./data/gse131013.rds")
setkey(topProbes, seqnames, start, end)
gse131013CpGs <- foverlaps(gse131013,
                           topProbes, 
                           by.x=c("seqnames", "start", "end"),
                           by.y=c("seqnames", "start", "end"),
                           type=c("any"),
                           nomatch=NULL)

gse131013CpGs[,condition:=tstrsplit(sample, "_", keep=2)]
gse131013CpGs[,cond_bin:=fifelse(condition=="T","T", "NM")]

# Get per probe x sample mean methylation value
gse131013Probes <- gse131013CpGs[,.(Beta=mean(Beta, na.rm=TRUE)),
                                    by=c("sample", "probe_id", "cond_bin")]

# Get per sample mean methylation values
gse131013Sample <- gse131013CpGs[,.(Beta=mean(Beta, na.rm=TRUE)),
                                    by=c("sample", "cond_bin")]
gse131013Sample <- subset(gse131013Sample, !is.na(Beta))
saveRDS(gse131013Sample, "./data/gse131013SampleTopProbes.rds")
```

## GSE48684

Download data and reshape:
```{r, load GSE48684 data}
for (i in 1:10) gc()
hm450.hg19 <- get450k(genome='hg19')
probeIds <- names(hm450.hg19)
hm450.hg19 <- as.data.table(hm450.hg19)
hm450.hg19$probe_Id <- probeIds

# load series and platform data from GEO
gset <- getGEO("GSE48684", GSEMatrix =TRUE, getGPL=FALSE)
gse48684 <- as.data.table(exprs(gset$GSE48684_series_matrix.txt.gz), keep.rownames=TRUE)
annotGSE48684 <- as.data.table(gset$GSE48684_series_matrix.txt.gz@phenoData@data)

gse48684 <- melt(gse48684, id.vars=c("rn"))
setnames(gse48684, c("rn", "variable", "value"), c("target_id","sample", "Beta"))

gse48684 <- merge(gse48684, 
                  hm450.hg19[,c("seqnames", "start", "end", "probe_Id"), with=FALSE],
                  by.x=c("target_id"),
                  by.y=c("probe_Id"),
                  all.x=FALSE, all.y=FALSE)

gse48684 <- merge(gse48684, 
                  annotGSE48684[,c("geo_accession", "disease status:ch1"), with=FALSE],
                  by.x=c("sample"),
                  by.y=c("geo_accession"))

setnames(gse48684, c("disease status:ch1"), c("condition"))
saveRDS(gse48684, "./data/gse48684.rds")
```

Get methylation rates of probes: 
```{r, methylation level probes GSE48684}
for (i in 1:10) gc()

# Overlap with top probes
setkey(topProbes, seqnames, start, end)
gse48684CpGs <- foverlaps(gse48684,
                          topProbes, 
                          by.x=c("seqnames", "start", "end"),
                          by.y=c("seqnames", "start", "end"),
                          type=c("any"),
                          nomatch=NULL)

for (i in 1:10) gc()

# label (tumor vs non-mucosa)
gse48684CpGs[,cond_bin:=fifelse(grepl("normal",condition),"NM", "T")]

# Get per probe x sample mean methylation value
gse48684Probes <- gse48684CpGs[,.(Beta=mean(Beta, na.rm=TRUE)),
                                by=c("sample", "cond_bin", "probe_id")] # , "disease status:ch1", probe_id

# Get per sample mean methylation values
gse48684Sample <- gse48684CpGs[,.(Beta=mean(Beta, na.rm=TRUE)),
                                by=c("sample", "cond_bin")] # , "disease status:ch1", probe_id
gse48684Sample <- subset(gse48684Sample, !is.na(Beta))
saveRDS(gse48684Sample, "./data/gse48684SampleTopProbes.rds")
```

## Training set 
```{r, load training set}
bismarkBSseqTrain <- readRDS("./data/bismarkBSseq_train_ids.rds")
topProbesRanges <- makeGRangesFromDataFrame(as.data.frame(topProbes))
seqlevelsStyle(topProbesRanges) <- "NCBI"

# Get the methylated fraction per probe
metTrain <- bsseq::getMeth(bismarkBSseqTrain, regions=topProbesRanges, 
                    type="raw", what="perRegion")
rownames(metTrain) <- topProbes$probe_id

# reshape the data & annotate
metTrain <- as.data.table(metTrain, keep.rownames="probe_id")
colnames(metTrain) <- c("probe_id", as.character(pData(bismarkBSseqTrain)$sample))

metTrain <- melt(metTrain, id.vars=c("probe_id"), 
                           variable.name="sample",
                           value.name="met_frac")

# label the samples
metTrain[,c("patient", "type"):=tstrsplit(sample, "_")]
metTrain[,cond_bin:=fifelse(type=="NM", "NM", "T")]

# Get per sample mean methylation values
metTrainSample <- metTrain[,.(met_frac=mean(met_frac, na.rm=TRUE)), 
                              by=c("sample", "cond_bin")]

metTrainSample <- subset(metTrainSample, !is.na(met_frac))
```

## Test set

```{r, load test set}
bismarkBSseqTest <- readRDS("./data/bismarkBSseq_test_ids.rds")
topProbesRanges <- makeGRangesFromDataFrame(as.data.frame(topProbes))
seqlevelsStyle(topProbesRanges) <- "NCBI"

# Get the methylated fraction per probe
metTest <- bsseq::getMeth(bismarkBSseqTest, regions=topProbesRanges, 
                   type="raw", what="perRegion")
rownames(metTest) <- topProbes$probe_id

# reshape the data & annotate
metTest <- as.data.table(metTest, keep.rownames="probe_id")
colnames(metTest) <- c("probe_id", as.character(pData(bismarkBSseqTest)$sample))

metTest <- melt(metTest, id.vars=c("probe_id"), 
                         variable.name="sample",
                         value.name="met_frac")

# label the samples
metTest[,c("patient", "type"):=tstrsplit(sample, "_")]
metTest[,cond_bin:=fifelse(type=="NM", "NM", "T")]

# Get per sample mean methylation values
metTestSample <- metTest[,.(met_frac=mean(met_frac, na.rm=TRUE)), by=c("sample", "cond_bin")]
metTestSample <- subset(metTestSample, !is.na(met_frac))
```

## Stratification by probes

```{r, roc}
gse48684Sample$dataset <- "GSE48684"
gse131013Sample$dataset <- "GSE131013"
cols <- c("sample", "Beta", "cond_bin", "dataset")

extDataSamples<- rbind(gse48684Sample[,cols,with=FALSE], 
                       gse131013Sample[,cols,with=FALSE])

extDataRoc <- getRocs(extDataSamples, scores="Beta", models="dataset",
                      labels="cond_bin", posClass="T", negClass="NM", 
                      subSample=FALSE)

metTestSample$dataset <- "test set"
metTestRoc <- getRocs(metTestSample, scores="met_frac", models="dataset",
                      labels="cond_bin", posClass="T", negClass="NM",
                      subSample=FALSE)

rocs <- rbindlist(list(extDataRoc, metTestRoc), use.names=TRUE, fill=TRUE)
aucDatasets <- rocs[,.(auc=unique(auc_mod)),
                       by=c("dataset")]

rocPlot <- ggplot(rocs,  aes(fpr, tpr, colour=dataset))+
  #scale_x_sqrt()+ 
  scale_color_brewer(palette="Dark2")+
  geom_text(x=0.75, 
                   y=c(0.1,0.3,0.5), 
           data=aucDatasets, 
           aes(colour=dataset, label=paste("AUC:", round(auc, 3))))+
  geom_line(aes(group=dataset), alpha=0.7, size=1.7)+
  labs(title="Validation datasets",
       subtitle="top DM Probes",
       x="False Positive Rate",
       y="True Positive Rate")+
  theme_ipsum(axis_title_size=12)
rocPlot

ggsave(file="./plots/rocPlotTest.png", rocPlot, width=20, height=15, units=c("cm"))
```

Heatmap of the top probes (for the matched samples)
```{r, top probes heatmap}
# Annotate (matched) datasets
gse131013Probes$dataset <- "GSE131013"
metTest$dataset <- "test set"
metTrain$dataset <- "train set"

# Combine
setnames(gse131013Probes, c("Beta"), c("met"))
setnames(metTest, c("met_frac"), c("met"))
setnames(metTrain, c("met_frac"), c("met"))

cols <- c("probe_id", "sample", "met", "dataset", "cond_bin")
allMetProbes <- rbindlist(list(gse131013Probes, 
                               metTrain[,cols,with=FALSE],
                               metTest[,cols, with=FALSE]), 
                          use.names=TRUE)

allMetProbes[,patient:=tstrsplit(sample, split="_", keep=1)]

# Calculate Methylation differences
allMetProbesCancer <- subset(allMetProbes, cond_bin=="T")
allMetProbesNormal <- subset(allMetProbes, cond_bin=="NM")

allMetProbesDiff <- merge(allMetProbesCancer,
                          allMetProbesNormal,
                          by=c("patient", "probe_id", "dataset"),
                          suffixes=c(".cancer", ".normal"))
allMetProbesDiff[,diff:=met.cancer-met.normal]

# order Patients
meanDiffPatient <- allMetProbesDiff[,.(mean_methylation_patient=mean(diff, na.rm=TRUE)), 
                                       by=c("patient")]$mean_methylation_patient
ordPatients <- order(meanDiffPatient)
patients <- unique(allMetProbesDiff$patient)
allMetProbesDiff$patient <- factor(allMetProbesDiff$patient, 
                                   levels=patients[ordPatients])

# order probes
allMetProbesDiff$probe_id <- factor(allMetProbesDiff$probe_id, 
                                    levels=topProbes$probe_id)

# order datasets
allMetProbesDiff$dataset <- factor(allMetProbesDiff$dataset, 
                                   levels=c("train set", "test set", 
                                            "GSE131013"))
allMetProbesDiff$probe_type <- "single probes"

# add annotation layer with global methylation difference 
# the probe panel (i.e. per patient)
allMetSamplesDiff <- allMetProbesDiff[,.(mean_met_diff=mean(diff, na.rm=TRUE)),
                                         by=c("patient", "dataset")]
allMetSamplesDiff$probe_id <- "all"
allMetSamplesDiff$probe_type <- "panel"
allMetSamplesDiffNeg <- subset(allMetSamplesDiff, mean_met_diff<=0)

heatmapProbes <- ggplot(allMetProbesDiff, aes(patient, probe_id, fill=diff)) + 
                      geom_tile(colour = "black", size=0.01)+
    scale_fill_viridis(option="inferno",
                       alpha=0.9,
                       guide=guide_colourbar(title="Methylation difference Probe"))+
    facet_grid(cols=vars(dataset), scales="free", 
               rows=vars(probe_type), space="free")+
    scale_x_discrete(position = "bottom")+
    ylab("Top 30 probes")+
    geom_point(data=allMetSamplesDiff, 
               size=2,
               aes(x=patient, y=probe_id, 
                   fill=mean_met_diff, 
                   color=mean_met_diff))+
    geom_tile(data=allMetSamplesDiffNeg, 
              color="darkred", fill=NA, na.value = NA, alpha=0, size=0.5)+
    scale_color_viridis(option="viridis",
                       alpha=0.9,
                       guide=guide_colourbar(title="Methylation difference Panel"))+
    theme_ipsum()+
    theme(axis.text.y = element_blank(),
          axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          panel.spacing = unit(0.5, "lines"),
          strip.background.x=element_rect(color="black", fill=NA, linetype="solid"),
          #strip.background.y=element_rect(color="black", fill=NA, linetype="solid"),
          #axis.title.y =element_blank(),
          strip.text.y = element_text(angle = 360, face="bold"),
          strip.text.x=element_text(face="bold"))
heatmapProbes

ggsave(file="./plots/heatmapTopProbesTotal.svg", 
       heatmapProbes, width=40, height=15, units=c("cm"))
```
Paired-sample ROC curves (?)

Show direction of methylation of the top probes
```{r, methylation change direction plot}
setnames(metTrainSample, "met_frac", "met")
setnames(metTestSample, "met_frac", "met")
setnames(gse131013Sample, "Beta", "met")

# Combine
metTrainSample$dataset <- "train set"
allPairedSamples <- rbindlist(list(metTrainSample,
                                   metTestSample,
                                   gse131013Sample), use.names=TRUE)

# Annotate
allPairedSamples[,patient:=tstrsplit(sample, split="_", keep=1)]
allPairedSamples[,cond_bin:=fifelse(cond_bin=="NM", "normal", "adenoma")]

# Define order
allPairedSamples <- merge(allPairedSamples,
                          allMetSamplesDiff[,c("patient", "mean_met_diff"), with=FALSE],
                          by=c("patient"))
allPairedSamples$patient <- factor(allPairedSamples$patient,
                                   levels=patients[ordPatients])

# add ranges
allPairedSamples[,min_cond:=cond_bin[which.min(met)], by=c("patient")]
allPairedSamples[,min_met:=min(met), by=c("patient")]
allPairedSamples[,max_met:=max(met), by=c("patient")]
allPairedSamples[,type:=fifelse(min_cond=="normal", "hyper", "hypo")]

# Define colors to be plotted for conditions
colorsConds <- brewer.pal(n=3,name="Set2")[1:2]
conds <- c("normal", "adenoma")
names(colorsConds) <- conds

# Define colors to be plotted for ranges
colorDirection <- c("lightgray", "darkred")
directions <- c("hyper", "hypo")
names(colorDirection) <- directions

# Define order of datasets
allPairedSamples$dataset <- factor(allPairedSamples$dataset,
                                   levels=c("train set", "test set", "GSE131013"))

dirPlot <- ggplot(allPairedSamples, aes(x=patient, y=met, fill=cond_bin))+
           facet_grid(cols=vars(dataset), scale="free_x", space="free_x")+
           geom_linerange(aes(ymin=min_met,ymax=max_met, color=type), size=0.5)+
           geom_point(pch=21, color="black")+
           scale_fill_manual(values=colorsConds, breaks=names(colorsConds),
                             guide=guide_legend(title="condition"))+
           scale_color_manual(values=colorDirection, breaks=names(colorDirection),
                              guide=guide_legend(title="direction"))+
           #scale_size_manual(values=alphaValues)+
           labs(x="individual",
                y="methylation (panel)")+
            theme_ipsum()+
            theme(axis.text.x=element_blank(),
                  axis.ticks.x=element_blank(),
                  panel.grid.major.x=element_blank())
dirPlot

ggsave(file="./plots/topProbesDirections.svg", 
       dirPlot, width=30, height=15, units=c("cm"))
ggsave(file="./plots/topProbesDirections.png", 
       dirPlot, width=30, height=15, units=c("cm"))
```

```{r, combine probes plot}
library(patchwork)
patchworkPlot = heatmapProbes/dirPlot
patchworkPlot[[2]] = patchworkPlot[[2]] + theme(strip.text.x = element_blank())
patchworkPlot

ggsave(file="./plots/topProbesFull.svg", 
       patchworkPlot, width=40, height=25, units=c("cm"))
ggsave(file="./plots/topProbesFull.png", 
       patchworkPlot, width=40, height=25, units=c("cm"))
```
Relative risk scores: 
```{r, relative risk scores}
sampleInfo <- colData(bismarkBSseqTest)
sampleInfo <- as.data.table(sampleInfo)
typeCounts <- sampleInfo[,.(n_type=.N), by=c("condition")]

# relative per probe risk scores
metTestPl <- subset(metTest, type!="NM")
metTestNM <- subset(metTest, type=="NM")
cols <- c("patient", "probe_id", "met")

metTestPatient <- merge(metTestPl[, cols, with=FALSE], 
                        metTestNM[, cols, with=FALSE], 
                        by=c("patient", "probe_id"))
setnames(metTestPatient, c("met.x", "met.y"), c("met_pl", "met_nm"))
metTestPatient[,relative_risk:=met_pl/met_nm]

min(metTestPatient$relative_risk, na.rm=TRUE)
max(metTestPatient$relative_risk, na.rm=TRUE)
```

# Probe ranking

```{r, ordering probes}
resultsProbes <- resultsProbes[order(fdr),]
write.table(resultsProbes, file="./data/probeRanking.csv", quote=FALSE,
            sep=",", eol="\r", row.names=FALSE)
```

Comparison to cologuard probes
```{r, ranking probes}
resultsProbes$rank <- 1:nrow(resultsProbes)

# rank cologuard markers
# probe_id %in% c(944, 383)
cologuardRanks <- subset(resultsProbes, grepl("944", master_id) | grepl("383", master_id))
write.table(cologuardRanks, file="./data/cologuardRanks.csv", quote=FALSE,
            sep=",", eol="\r", row.names=FALSE)
knitr::kable(cologuardRanks, "simple")
```

# CpG Ranking

```{r, ranking cpgs}
resultsCpG <- subset(resultsCpG, method=="Voom+Limma")
resultsCpG <- resultsCpG[order(adj_p_values),]

# add CpG coordinate from illumina
write.table(resultsCpG, file="./data/CpGRanking.csv", quote=FALSE,
            sep=",", eol="\r", row.names=FALSE)
```

# Session Information

```{r, sessioninfo}
sessionInfo()
```
